Two Decades of Smalltalk VM Development

### 2018-08-17 paper.tex
Add a placeholder for Dan.
fixed references according to my maniac format. Added ref to old VMIL paper. Added Dan as a co-author

\usepackage{xcolor}
\newcommand{\todo}[1]{\color{orange}\fbox{\bfseries\sffamily\scriptsize TODO:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
\newcommand{\sd}[1]{\color{red}\fbox{\bfseries\sffamily\scriptsize Stef:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
\newcommand{\egb}[1]{\color{blue}\fbox{\bfseries\sffamily\scriptsize Elisa:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
\newcommand{\eem}[1]{\color{green}\fbox{\bfseries\sffamily\scriptsize Eliot:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
\newcommand{\cba}[1]{\color{purple}\fbox{\bfseries\sffamily\scriptsize Clement:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
%\newcommand*{rotatebox{75}}
\input{macros}

@@

@@

Over the years, the Squeak VM evolved to give birth recently to \openSmalltalkVM\footnote{https://github.com/OpenSmalltalk/opensmalltalk-vm/}, the default VM for various Smalltalk-like systems such as Pharo \cite{PharoByExample}, Squeak \cite{SqueakByExample}, Cuis, Croquet and Newspeak \cite{NewspeakOopsla}. 
As the VM evolved, the original simulator co-evolved as a tool to develop and debug the VM. The meta data used to guide the translation process was replaced by \emph{pragmas} \cite{Pragmas}.  
%One change moved the meta data that guides translation out of the executable Smalltalk into pragmas \cite{Pragmas}.  
But the most significant evolution of the simulator came with the introduction of the Just-In-Time compiler (JIT). 
%The code base was liberally sprinkled with assertions. 
The spaghetti stack was mapped to a more conventional stack frame organisation in a stack zone of a few hundred k bytes organised into small pages, a scheme called context-to-stack mapping \cite{EfficientST80}, \cite{StackMappingVW}.  Machine code generated by the JIT was executed by binding multiple processor simulators (first Bochs \cite{Bochs} for x86, SkyEye for ARMv6, Smalltalk code for 32-bit MIPS).  
Finally, a new object representation and garbage collector was added to improve performance and also to support 64-bits \cite{Spur}.  This required refactoring the interpreter to allow the object representation to be chosen at startup \footnote{The JIT was written with this eventuality in mind.}. Bochs was used a second time for x64 support.

>>>>

\usepackage{xcolor}
\newcommand{\todo}[1]{\color{orange}\fbox{\bfseries\sffamily\scriptsize TODO:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
\newcommand{\egb}[1]{\color{blue}\fbox{\bfseries\sffamily\scriptsize Elisa:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
\newcommand{\eem}[1]{\color{green}\fbox{\bfseries\sffamily\scriptsize Eliot:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
\newcommand{\cba}[1]{\color{purple}\fbox{\bfseries\sffamily\scriptsize Clement:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
%\newcommand*{rotatebox{75}}
\input{macros}

@@

%% Author with two affiliations and emails.
\author{Daniel Ingalls}
\affiliation{
  % \position{}
	%\department{Software Languages Lab}              %% \department is recommended
	\institution{Y Combinator Research}            %% \institution is required
	\city{San Francisco}
  % \state{}
  % \postcode{}
	\country{California}                    %% \country is recommended
}
\email{TOFIX danhhingalls@gmail.com}          %% \email is recommended

@@

Over the years, the Squeak VM evolved to give birth recently to \openSmalltalkVM\footnote{https://github.com/OpenSmalltalk/opensmalltalk-vm/}, the default VM for various Smalltalk-like systems such as Pharo \cite{PharoByExample}, Squeak \cite{SqueakByExample}, Cuis, Croquet and Newspeak \cite{NewspeakOopsla}. 
As the VM evolved, the original simulator co-evolved as a tool to develop and debug the VM. The meta data used to guide the translation process was replaced by \emph{pragmas} \cite{Pragmas}.  
%One change moved the meta data that guides translation out of the executable Smalltalk into pragmas \cite{Pragmas}.  
But the most significant evolution of the simulator came with the introduction of the Just-In-Time compiler (JIT) \cite{CogJIT}. 
%The code base was liberally sprinkled with assertions. 
The spaghetti stack was mapped to a more conventional stack frame organisation in a stack zone of a few hundred k bytes organised into small pages, a scheme called context-to-stack mapping \cite{EfficientST80}, \cite{StackMappingVW}.  Machine code generated by the JIT was executed by binding multiple processor simulators (first Bochs \cite{Bochs} for x86, SkyEye for ARMv6, Smalltalk code for 32-bit MIPS).  
Finally, a new object representation and garbage collector was added to improve performance and also to support 64-bits \cite{Forwarders}.  This required refactoring the interpreter to allow the object representation to be chosen at startup\footnote{The JIT was written with this eventuality in mind.}. Bochs was used a second time for x64 support.
